openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding

### 2020-10-01 paper/policar2020-opentsne.tex
wide format

\documentclass[twocolumn]{article}

@@

\begin{abstract}
Point-based visualizations of large, multi-dimensional data from molecular
biology can reveal meaningful clusters. One of the most popular
techniques to construct such visualizations is $t$-distributed stochastic
neighbor embedding (t-SNE). Several extensions of t-SNE have
recently been proposed to address issues of scalability and the quality
of the resulting visualizations. We introduce \opentsne, a modular
Python library that implements the core t-SNE algorithm and its
extensions. The library is orders of magnitude faster than existing
popular implementations, including those from scikit-learn. Unique to
\opentsne\ is also the ability to map new data to existing embeddings,
which can surprisingly assist in solving batch effects.
\end{abstract}

@@

The abundance of high-dimensional data sets in molecular biology calls for
dimensionality reduction techniques, particularly for methods that produce
informative data visualizations. Popular approaches include principal component
analysis (PCA), multidimensional scaling, t-distributed stochastic neighbor
embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold
approximation and projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE
lately received much attention as it can address high volumes of data and reveal
a meaningful clustering structure. Increased resolution and throughput of modern
molecular assays has lead to the frequent use of t-SNE in diverse fields
including, but not limited to, single-cell
transcriptomics (scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
human genetics~\cite{hirata2019genetic}, metagenomic
assembly~\cite{beaulaurier2018metagenomic}, the spatial organization of
microbial communities~\cite{sheth2019spatial} and
metabolomics~\cite{tkachev2019differences}. Reports on single-cell gene
expression data, our running example, often start with an overview of the cell
landscape, where t-SNE embeds high-dimensional expression profiles into a
two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show
two such embeddings.

@@

Despite its utility, t-SNE has often been criticized for its limited
scalability, lack of global organization -- t-SNE emphasizes local clusters that
are arbitrarily scattered in the low-dimensional space -- and the absence of
theoretically-founded methods to map new data into existing
embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these
shortcomings have recently been addressed. Linderman \textit{et al.} sped-up the
method through an interpolation-based approximation, achieving linear complexity
in the number of samples~\cite{linderman2019fast}. Kobak \& Berens proposed
several techniques to improve global positioning, including estimating
similarities with a mixture of Gaussian kernels~\cite{kobak2019art}. In our
previous work, we presented a principled approach for embedding new samples into
existing visualizations~\cite{policar2019embedding}.

@@

\opentsne\ is compatible with the Python data science ecosystem and
libraries that include \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}),
thus providing a familiar and intuitive application program interface to the new
users. Its modular design encourages extensibility and experimentation with
various settings and changes in the analysis pipeline.  We have released
\opentsne\ in an open-source to support adaptations by the wider community. At
the time of the writing, the library averages thousands of weekly
downloads and has received almost 700 GitHub stars, an appreciation score gained
by only $0.3\%$ of public Python repositories. The source code is available at
\url{https://github.com/pavlin-policar/openTSNE} and supports installations from
\textsf{PyPI} and \textsf{conda-forge}, the two most widely adopted Python
package managers.

Accessibility of the latest theoretical advances in t-SNE is one of the core
design principles of \opentsne. These advances are accessible through a
programming interface of the library, where we closely follow the style of the
interface as implemented in \textsf{scikit-learn}~\cite{sklearn_api}, a popular
library for machine learning. The following snippet exemplifies this interface
and codes for four visualizations in Fig.~\ref{fig:tasic}, demonstrating the
effect of a particular recent theoretical advance in the t-SNE algorithm.

\begin{minted}[
fontsize=\footnotesize,
frame=lines,
framesep=2mm,
baselinestretch=1.1
]{python}
import openTSNE
from openTSNE.affinity import Multiscale
# a - standard t-SNE
tsne1 = openTSNE.TSNE(perplexity=30).fit(X)
# b - high perplexity  for globa structure
tsne2 = openTSNE.TSNE(perplexity=500).fit(X)
# c - multiscale kernel for local/global structure
aff = Multiscale(X, perplexities=[30, 500])
tsne3 = openTSNE.TSNE(affinities=aff).fit(X)
# d - decrease dof for higher resolution
tsne4 = openTSNE.TSNE(dof=0.6).fit(X)
\end{minted}

\noindent The code snippet considers a \textsf{numpy} array or \textsf{scipy}
sparse matrix \texttt{X} and creates four \texttt{TSNEEmbedding} objects. These
objects encode t-SNE embeddings that may subsequently be optimized under
different parameter settings or used to embed new samples into the embedding
landscape. Function calls may include additional parameters not shown in the
example code but explained in accompanying notebooks available at
\url{https://github.com/pavlin-policar/opentsne-paper}.

@@

Dimensionality reduction techniques implicitly assume that high-dimensional data
lies on a lower-dimensional manifold, which can accurately be captured by a
smaller number of dimensions. However, there is no evidence that every data set
can accurately be described by using only two dimensions, and any such embedding
will inevitably lead to a loss of information. Thus, it is beneficial to examine
multiple embeddings, each of which provides a different perspective on topology
and other data characteristics.

\begin{figure*}[htbp]
	\includegraphics[width=\textwidth]{tasic2018}
	\caption{\label{fig:tasic}We use \opentsne\ to create four different
	visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
	each providing a different perspective into the topology of the data.
	The data set contains 21,874 single-cells originating from the mouse
	neocortex. Cluster annotations and colors are taken from the original
	publication. Warm colors correspond to excitatory neurons, cool colors
	correspond to inhibitory neurons, and gray/brown colors correspond to
	non-neuronal cells. Standard t-SNE \textbf{(a)} emphasizes local
	structure while increasing perplexity \textbf{(b)} results in a more
	meaningful layout of the clusters. We can also combine the two
	perplexities by using a multiscale kernel affinity model \textbf{(c)}
	and obtain a trade-off between global and local structure.
	Alternatively, we can inspect more fine-grained structure and reveal
  smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}

We illustrate this point by generating four different embeddings of the data on
single-cell gene expression in mouse brain~\cite{tasic2018shared}.
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While
different clusters of excitatory and inhibitory neurons appear close to one
another, all clusters appear equidistant from their neighbors, and the overall
relations between groups are not obvious. The embedding in Fig.~\ref{fig:tasic}.b
focuses on preserving larger neighborhoods of points, resulting in a more
globally consistent layout where relations between clusters become more
apparent. Here, it is evident from the increased white space between groups that
there is one large class of excitatory neurons and two related classes of
inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods
leads to the absorption of smaller clusters into larger ones. Alternatively,
Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve
both the global organization of clusters and prevent the cluster absorption. We
constructed the embeddings from Fig.~\ref{fig:tasic}.d with the settings used
for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution. The distinction
between these two depictions reveals that many clusters are composed of
subgroups representing different cell subtypes.

@@

Exaggeration can highlight transitions between cell states in developmental
studies. Standard t-SNE often produces embeddings with clearly defined, discrete
clusters. We can adjust the level of granularity and resolution of the clusters
with several parameters in Fig.~\ref{fig:tasic}. However, discrete clusters are
often undesired in developmental studies where cells' state is changed
continuously. To this end, other embedding techniques such as UMAP and
ForceAtlas2~\cite{jacomy2014forceatlas2} tend to better capture the continuity
between cell states. Recently, B{\"o}hm \textit{et al.}~\cite{bohm2020unifying}
have shown that embeddings produced by t-SNE with exaggeration values of 4 and
$\sim30$ construct embeddings which are markedly similar to UMAP and
ForceAtlas2, respectively. For example, in Fig.~\ref{fig:cao}.a, the
developmental trajectory between different time points is difficult to observe
due to many sprawled out clusters. On the other hand, it is easier to trace the
development when we increase the exaggeration factor from $1$ to $4$
(Fig.~\ref{fig:cao}.b-c).

@@

Unlike other popular dimensionality reduction methods such as principal
component analysis or autoencoders, t-SNE is a non-parametric method and does
not define an explicit mapping to the embedding space. Therefore embeddings of
new data points need to be found through
optimization~\cite{policar2019embedding}. \opentsne\ is currently the only
publicly available library allowing users to add new samples to existing
embeddings.

Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a
previously labeled single-cell data set and embed cells from a separate
experiment into the reference landscape. The reference data from
Macosko~\textit{et al.}~\cite{macosko2015highly} contains gene expression
profiles from mouse retinal cells. By embedding the samples from a similar
experiment on bipolar retinal cells by Shekhar~\textit{et
al.}~\cite{shekhar2016comprehensive}, we can correctly map the bipolar cell
clusters in the reference.

Embedding single cells into existing reference atlases can also be useful for
cell-type classification in cases of unknown cell identities. For instance, in
Fig.~\ref{fig:transform}, we construct a reference embedding using labeled data
from Hochgerner~\textit{et al.}~\cite{hochgerner2018conserved} containing
gene-expression profiles of cells from the mouse brain. The authors assign a
type to each cell. We can verify their classification accuracy by visualizing
the expression of well-established gene markers for the major cell types. We
then embed cells from Harris~\textit{et al.}~\cite{harris2018classes} into a
constructed atlas. In Harris~\textit{et al.}~, labels are provided only for
neuronal cells. In the resulting mapping, we can quickly identify other
non-neuronal cell types, including oligodendrocytes and astrocytes. We can
further use marker genes to validate that the mapping in the reference landscape
was correct.

@@

The examples presented above demonstrate how to use \opentsne\ to quickly gain
insight into a newly-sequenced, single-cell data sets by utilizing existing cell
atlases. The approach is general and not limited to single-cell gene expression,
and one can, in principle, apply it to any tabular data set regardless of field.

@@

Versatility, the ability to use and combine different optimization approaches to
construct the embedding spaces, is another of \opentsne's core design
principles. Kobak \& Berens recently introduced several recommendations and
tricks to obtain better and more meaningful t-SNE
visualizations~\cite{kobak2019art}. These include multi-scale similarity
kernels, perplexity annealing, and increasing exaggeration when working with
more massive data sets. \opentsne\ provides a flexible program interface to
incorporate these improvements in just a few lines of code. Furthermore, \opentsne\
supports custom affinity models, enabling users to construct t-SNE embeddings on
non-tabular relational data: the only requirement imposed by the affinity-model
is the availability of similarities between data points.
Finally, \opentsne's comprehensive callback system can be utilized
to monitor and adapt different stages of the optimization phase and has been
used to construct visually appealing animations of the t-SNE optimization
process.

@@

One of the t-SNE's common criticisms is limited scalability to large data sets
containing, for instance, millions of data
points~\cite{becht2019dimensionality}. The culprit for slow response time stems
from a particular optimization procedure and its specific implementation in
popular Python libraries. Until quite recently, most popular
implementations of t-SNE were based on the Barnes-Hut approximation scheme
developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic
complexity $\mathcal{O}(N \log N)$, where $N$ is the number of data items ({\em
e.g.} cells). The most widely-used implementation of t-SNE came from
\textsf{scikit-learn}, which exhibits long runtimes when compared to its C++
counterpart -- \textsf{MulticoreTSNE}. The multi-threaded \textsf{MulticoreTSNE}
implementation~\cite{Ulyanov2016} can construct t-SNE embeddings of millions of
data points in a matter of hours on widely-accessible, consumer-grade processors
(Fig.~\ref{fig:benchmarks}). However, recently, Linderman \textit{et al.}
developed a new approximation scheme -- FIt-SNE -- which further reduces the
asymptotic time complexity to $\mathcal{O}(N)$. We have included this scheme in
\opentsne\, enabling the embedding of large data sets in a matter of minutes.

Fig.~\ref{fig:benchmarks} benchmarks four Python t-SNE implementations,
including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE}
(v0.1), \textsf{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.4.3). We perform
benchmarks on two computational platforms, one representing the usage on a
personal computer and the other utility of these libraries on a high-performance
computing platform. The Intel(R) Core i7 is commonly found in consumer-grade
laptop computers, while Intel(R) Xeon(R) processors appear in high-performance
computing machines. Benchmarks were run for $1,000$ iterations with the original
t-SNE parameters, as some implementations do not allow their modification.

@@

Intuitive access and simple installation procedures almost universally correlate
with the widespread adoption of novel computational techniques. While the t-SNE
from \textsf{scikit-learn} fits this requirement, its implementation is
prohibitively slow for even moderately-sized data sets that span tens of
thousands of data records. Other C++ implementations such as
\textsf{MulticoreTSNE} and \textsf{FIt-SNE} exhibit better scaling in more
massive data sets, but do not provide precompiled binaries and require users to
compile the software themselves. This problem is critical, for instance, for
users of Windows, where the C++ compiler does not come with the system, making
the correct configuration of current t-SNE implementations cumbersome.

We have designed \opentsne\ to make it accessible to a broader audience. We
provide precompiled binaries for all major Python versions on all major
platforms, making the installation process as seamless as possible. One can
install \opentsne\ through the Python Package Index (\textsf{PyPI}) or \textsf{conda} from the
\textsf{conda-forge} channel. \opentsne's interface is inspired by \textsf{scikit-learn},
which is well established in the Python data science ecosystem. \opentsne\
implements multi-threaded versions of both the Barnes-Hut and FIt-SNE
approximation schemes, enabling it to be applied to data sets containing millions of data
points. While the Python virtual machine inevitably introduces some performance overhead,
the runtime is comparable to its C++ counterpart -- \textsf{FIt-SNE}. Finally,
\opentsne\ is extensible. Its modular design enables researchers to quickly
experiment with different parameter settings and easily incorporate their own components into
the software. We provide a full feature list and comparison to other popular
t-SNE implementations in Table~\ref{tab:features}.

\begin{table}
\caption{\label{tab:features}
  We compare features of \opentsne\ to three popular open-source implementations
	from \textsf{scikit-learn}~(v0.23.1), \textsf{MulticoreTSNE} (v0.1), and
	\textsf{FIt-SNE} (v1.1.0).  The first section of the comparison
	addresses packaging and distribution. A properly packaged library is
	easily accessible to users, and developers should easily include it in
	dependency lists of other software packages. The second section of the
	comparison lists the two existing t-SNE approximation schemes. The
	FIt-SNE approximation scheme is required for t-SNE to scale up to
	millions of data points. The final section provides a list of extensions
	and improvements to the standard t-SNE algorithm, many of which can
	produce markedly better visualizations.
}

\newcommand*\rot{\rotatebox{90}}
\renewcommand{\arraystretch}{1.25}

\begin{tabular}{r c c c c c c}
\toprule
\setlength\tabcolsep{6pt}
& \rot{\textsf{scikit-learn}} & \rot{\textsf{MulticoreTSNE}} & \rot{\textsf{FIt-SNE}} & \rot{\textsf{openTSNE}} \\
\toprule
\textsf{PyPI} package & \checkmark & \checkmark & & \checkmark \\
\textsf{conda} package & \checkmark & & & \checkmark \\
Precompiled binaries & \checkmark & & & \checkmark \\
\hline
Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & & \checkmark \\
FIt-SNE ($\mathcal{O}(N)$) & & & \checkmark & \checkmark \\
\hline
Multiscale Gaussian kernels & & & \checkmark & \checkmark \\
Fully-custom affinity kernels & & & & \checkmark \\
Variable degrees of freedom & & & \checkmark & \checkmark \\
Variable exaggeration & & & \checkmark & \checkmark \\
Better initialization & & & \checkmark & \checkmark \\
Automatic learning rate & & & \checkmark & \checkmark \\
Embedding new samples & & & & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

>>>>

% \documentclass[twocolumn]{article}
\documentclass[letter]{article}
\usepackage[a4paper, total={6in, 8in}]{geometry}

@@

\begin{abstract}
Point-based visualizations of large, multi-dimensional data from molecular
biology are often used to reveal meaningful clusters. One of the most popular
techniques to construct such visualizations is $t$-distributed stochastic
neighbor embedding (t-SNE). Several extensions of t-SNE have
recently been proposed to address issues of scalability and the quality
of the resulting visualizations. We introduce \opentsne, a modular
Python library that implements the core t-SNE algorithm and its many
extensions.
The library is fast and enables users to embed data sets containing millions of
data points in a matter of minutes on consumer-grade laptop computers.
Unique to \opentsne\ is also the ability to map new data to existing embeddings,
which can surprisingly assist in solving batch effects.
\end{abstract}

@@

The abundance of high-dimensional data sets in molecular biology calls for
dimensionality reduction techniques that are able to produce
informative data visualizations. Popular approaches include principal component
analysis (PCA), multidimensional scaling, t-distributed stochastic neighbor
embedding (t-SNE)~\cite{maaten2008visualizing}, and uniform manifold
approximation and projections (UMAP)~\cite{2018arXivUMAP}. Among these, t-SNE has
received much attention as it can address high volumes of data and reveal
meaningful clustering structure. The increased resolution and throughput of modern
molecular assays has lead to the frequent use of t-SNE in diverse fields
including, but not limited to, single-cell
transcriptomics (scRNA-seq,~\cite{macosko2015highly,cao2019single,tasic2018shared}),
human genetics~\cite{hirata2019genetic}, metagenomic
assembly~\cite{beaulaurier2018metagenomic}, the spatial organization of
microbial communities~\cite{sheth2019spatial} and
metabolomics~\cite{tkachev2019differences}. Reports on single-cell gene
expression data, our running example, often start with an overview of the cell
landscape, where t-SNE is used to embed high-dimensional expression profiles into a
two-dimensional space. Figs.~\ref{fig:macosko}.a and \ref{fig:macosko}.b show
two such embeddings.

@@

Despite its utility, t-SNE has often been criticized for its limited
scalability, lack of global organization -- t-SNE identifies well-defined clusters that
may be arbitrarily scattered throguhout the low-dimensional space -- and the absence of
theoretically-founded methods to map new data into existing
embeddings~\cite{ding2018interpretable,becht2019dimensionality}. Most of these
shortcomings have recently been addressed. Linderman \textit{et al.} developed
an interpolation-based approximation scheme which massively improved the
scalability of t-SNE, achieving linear time complexity
in the number of samples~\cite{linderman2019fast}. Kobak \& Berens proposed
several techniques to improve global cluster coherence, including estimating
similarities with a mixture of Gaussian kernels~\cite{kobak2019art}. In our
previous work, we presented a principled approach for embedding new samples into
existing visualizations~\cite{policar2019embedding}.

@@

\opentsne\ is compatible with the Python data science ecosystem and
libraries including \textsf{numpy}, \textsf{scikit-learn}, \textsf{scanpy}),
providing a familiar and intuitive application program interface to new
users. Its modular design encourages extensibility and experimentation with
various settings and changes to the analysis pipeline. We have released
\opentsne\ to the open-source community to make recent theoretical advancements
more accessible to a wider audience. At
the time of the writing, the library averages thousands of weekly
downloads and has received almost 700 GitHub stars, an appreciation score gained
by only $0.3\%$ of public Python repositories. The source code is available at
\url{https://github.com/pavlin-policar/openTSNE} and supports installations from
\textsf{PyPI} and \textsf{conda-forge}, the two most widely adopted Python
package managers.

Accessibility of the latest theoretical advancments in t-SNE is one of the core
design principles of \opentsne. This range of advancments is easily accessible through an intuitive
programming interface, where we closely follow the style
\textsf{scikit-learn}~\cite{sklearn_api}, a popular and widely used
library for machine learning. The following snippet exemplifies this interface
and was used to generate the four visualizations in Fig.~\ref{fig:tasic}, demonstrating the
effect of a particular recent theoretical advancments in the t-SNE algorithm.

\begin{minted}[
fontsize=\footnotesize,
frame=lines,
framesep=2mm,
baselinestretch=1.1
]{python}
import openTSNE
from openTSNE.affinity import Multiscale
# a - standard t-SNE
tsne1 = openTSNE.TSNE(perplexity=30).fit(X)
# b - high perplexity for global structure
tsne2 = openTSNE.TSNE(perplexity=500).fit(X)
# c - multiscale kernel for local/global structure
aff = Multiscale(X, perplexities=[30, 500])
tsne3 = openTSNE.TSNE(affinities=aff).fit(X)
# d - decrease dof for higher resolution
tsne4 = openTSNE.TSNE(dof=0.6).fit(X)
\end{minted}

\noindent The code snippet considers a \textsf{numpy} array or \textsf{scipy}
sparse matrix \texttt{X} and creates four \texttt{TSNEEmbedding} objects. These
objects represent t-SNE embeddings that may subsequently be optimized under
different parameter settings or used to embed new samples into the embedding
landscape. Function calls may include additional parameters not shown in the
example code but are explained in the accompanying notebooks available at
\url{https://github.com/pavlin-policar/opentsne-paper}.

@@

Dimensionality reduction techniques implicitly assume that high-dimensional data
lies on a lower-dimensional manifold, which can accurately be captured by a
small number of dimensions. However, there is no evidence that every data set
can accurately be described using only two dimensions, and any such embedding
will inevitably lead to a loss of information. Thus, it is beneficial to examine
multiple embeddings, each of which provides a different perspective on topology
and other data characteristics.

\begin{figure*}[htbp]
	\includegraphics[width=\textwidth]{tasic2018}
	\caption{\label{fig:tasic}
  We use \opentsne\ to create four different
	visualizations of the Tasic \textit{et al.}~\cite{tasic2018shared} data,
	each providing a different perspective into the topology of the data.
	The data set contains 21,874 single-cells originating from the mouse
	neocortex. Cluster annotations and colors are taken from the original
	publication. Warm colors correspond to excitatory neurons, cool colors
	correspond to inhibitory neurons, and gray/brown colors correspond to
	non-neuronal cells. Standard t-SNE \textbf{(a)} emphasizes local
	structure while increasing perplexity \textbf{(b)} results in a more
	meaningful layout of the clusters. We can also combine the two
	perplexities by using a multiscale kernel affinity model \textbf{(c)}
	and obtain a trade-off between global and local structure.
	Alternatively, we can inspect more fine-grained structure and reveal
  smaller clusters by using a more heavy-tailed kernel \textbf{(d)}.}
\end{figure*}

We illustrate this point by generating four different embeddings of the data on
single-cell gene expression in mouse brain~\cite{tasic2018shared}.
Fig.~\ref{fig:tasic}.a shows an embedding using default t-SNE parameters. While
different clusters of excitatory and inhibitory neurons appear close to one
another, all clusters appear equidistant from their neighbors, and the overall
relations between groups are not obvious. The embedding in Fig.~\ref{fig:tasic}.b
focuses on preserving larger neighborhoods of points, resulting in a more
globally consistent layout where relations between clusters become more
apparent. Here, it is evident from the increased white space between groups that
there is one large class of excitatory neurons and two related classes of
inhibitory neurons. Unfortunately, focusing on preserving large neighborhoods
leads to the absorption of smaller clusters into larger ones. Alternatively,
Fig.~\ref{fig:tasic}.c uses multi-scale similarity kernels that aim to preserve
both the global organization of clusters and prevent smaller cluster absorption. We
constructed the embedding from Fig.~\ref{fig:tasic}.d with the settings used
for Fig.~\ref{fig:tasic}.a, but at a finer level of resolution.
The figure demonstrates that some clusters are composed of numerous, smaller
subgroups representing different cell subpopulations which are not visible under
standard parameter settings.

@@

Exaggeration can highlight transitions between cell states in developmental
studies. Standard t-SNE often produces embeddings with clearly defined, discrete
clusters. We can adjust the level of granularity and resolution of the clusters
with several parameters in Fig.~\ref{fig:tasic}. However, discrete clusters are
often undesired in developmental studies where cells' state is assumed to follow
a continuous transition path. To this end, other embedding techniques such as UMAP and
ForceAtlas2~\cite{jacomy2014forceatlas2} are used to better capture the continuity
between cell states. Recently, B{\"o}hm \textit{et al.}~\cite{bohm2020unifying}
showed that embeddings produced by t-SNE with exaggeration values of 4 and
$\sim30$ construct embeddings which are markedly similar to UMAP and
ForceAtlas2, respectively. For example, in Fig.~\ref{fig:cao}.a, the
developmental trajectory between different time points is difficult to observe
due to many sprawled out clusters. On the other hand, it is easier to trace the
development when we increase the exaggeration factor from $1$ to $2$ to $4$ in
Figs.~\ref{fig:cao}.b-c.

@@

Unlike other popular dimensionality reduction techniques such as principal
component analysis or autoencoders, t-SNE is a non-parametric method and does
not define an explicit mapping to the embedding space. Therefore embeddings of
new data points need to be found through
optimization~\cite{policar2019embedding}. \opentsne\ is currently the only
publicly available library allowing users to add new samples to existing
embeddings in a principled manner.

Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a
previously labeled single-cell data set and embed cells from a separate
experiment into the reference landscape. The reference data from
Macosko~\textit{et al.}~\cite{macosko2015highly} contains gene expression
profiles from mouse retinal cells. By embedding the samples from a similar
experiment on bipolar retinal cells by Shekhar~\textit{et
al.}~\cite{shekhar2016comprehensive}, we can correctly map the bipolar cell
clusters onto the reference embedding.

Embedding single cells into existing reference atlases can also be useful for
cell-type classification in cases of unknown cell identities. For instance, in
Fig.~\ref{fig:transform}, we construct a reference embedding using labeled data
from Hochgerner~\textit{et al.}~\cite{hochgerner2018conserved} containing
gene-expression profiles of cells from the mouse brain. The authors assign a
type to each cell. We can verify their classification accuracy by visualizing
the expression of well-established gene markers for the major cell types. We
then embed cells from Harris~\textit{et al.}~\cite{harris2018classes} into the
constructed cell atlas. In Harris~\textit{et al.}~, labels are provided only for
neuronal cells. In the resulting mapping, we can quickly identify other
non-neuronal cell types, including oligodendrocytes and astrocytes. We can
further use marker genes to validate that the mapping in the reference landscape
is correct.

@@

The examples presented above demonstrate how to use \opentsne\ to quickly gain
insight into newly-sequenced, single-cell data sets by utilizing existing cell
atlases. The approach is general and not limited to single-cell gene expression,
and one can, in principle, apply it to any tabular data set regardless of field.

@@

Versatility, the ability to use and combine different optimization approaches to
construct different embedding spaces, is another of \opentsne's core design
principles. Kobak \& Berens recently provided several recommendations and
tricks to obtain better and more meaningful t-SNE
visualizations~\cite{kobak2019art}. These include multi-scale similarity
kernels, perplexity annealing, and increasing exaggeration when working with
massive data sets. \opentsne\ provides a flexible program interface to
incorporate these improvements in just a few lines of code. Furthermore, \opentsne\
supports custom affinity models, enabling users to construct t-SNE embeddings on
non-tabular relational data: the only requirement imposed by the affinity-model
is some notion of similarity between data points.
Finally, \opentsne's comprehensive callback system can be utilized
to monitor and adapt different stages of the optimization phase and has been
used to construct visually appealing animations of the t-SNE optimization
process.

@@

One of the t-SNE's common criticisms is limited scalability to large data sets
containing, for instance, millions of data
points~\cite{becht2019dimensionality}. The culprit for slow response time stems
from a particular optimization procedure and its specific implementation in
popular Python libraries. Until quite recently, most popular
implementations of t-SNE were based on the Barnes-Hut approximation scheme
developed by van der Maaten in 2014~\cite{van2014accelerating} with asymptotic
time complexity $\mathcal{O}(N \log N)$, where $N$ is the number of data items ({\em
e.g.} cells). The most widely-used implementation of t-SNE came from
\textsf{scikit-learn}, which exhibits long runtimes when compared to its C++
counterpart -- \textsf{MulticoreTSNE}. The multi-threaded \textsf{MulticoreTSNE}
implementation~\cite{Ulyanov2016} can construct t-SNE embeddings of millions of
data points in a matter of hours on widely-accessible, consumer-grade processors
(Fig.~\ref{fig:benchmarks}). However, recently, Linderman \textit{et al.}
developed a new approximation scheme -- FIt-SNE -- which further reduces the
asymptotic time complexity to $\mathcal{O}(N)$. We include this approach in
\opentsne\, enabling the embedding of large data sets in a matter of minutes.

Fig.~\ref{fig:benchmarks} benchmarks four popular Python t-SNE implementations,
including those from \textsf{scikit-learn} (v0.23.1), \textsf{MulticoreTSNE}
(v0.1), \textsf{FIt-SNE} (v1.1.0), and our \opentsne\ (v0.4.3). We perform
benchmarks on two computational platforms, one representing the usage on a
personal computer and the other the utility of these libraries on a high-performance
computing platform. The Intel(R) Core i7 is commonly found in consumer-grade
laptop computers, while Intel(R) Xeon(R) processors appear in high-performance
computing machines. Benchmarks were run for $1,000$ iterations with the original
t-SNE parameters, as some implementations do not allow their modification.

@@

Intuitive access and simple installation procedures almost universally correlate
with the widespread adoption of novel computational techniques. While the t-SNE implementation
from \textsf{scikit-learn} fits this requirement, the implementation proves
prohibitively slow for even moderately-sized data sets that span tens of
thousands of data records. Other C++ implementations such as
\textsf{MulticoreTSNE} and \textsf{FIt-SNE} exhibit better scaling in more
massive data sets, but do not provide precompiled binaries and require users to
compile the software themselves. This problem is critical, for instance, for
users of the Windows operating system, where the C++ compiler does not come with the system, making
the correct configuration of current t-SNE implementations cumbersome.

We designed \opentsne\ to be accessible to a broader audience. We
provide precompiled binaries for all major Python versions on all major
platforms, making the installation process as seamless as possible. One can
install \opentsne\ through the Python Package Index (\textsf{PyPI}) or \textsf{conda} from the
\textsf{conda-forge} channel, the two most widely adopted Python package managers.
\opentsne's interface is inspired by \textsf{scikit-learn},
which is well established in the Python data science ecosystem. \opentsne\
implements multi-threaded versions of both the Barnes-Hut and FIt-SNE
approximation schemes, enabling it to be applied to data sets containing millions of data
points. While the Python virtual machine inevitably introduces some performance overhead,
the runtime is comparable to its C++ counterpart -- \textsf{FIt-SNE}. Finally,
\opentsne\ is extensible. Its modular design enables researchers to quickly
experiment with different parameter settings and easily incorporate custom components into
the software. We provide a full feature list and comparison to other popular
t-SNE implementations in Table~\ref{tab:features}.

\begin{table}
\caption{\label{tab:features}
  Features of \opentsne\ compared to other three popular open-source implementations
	from \textsf{scikit-learn}~(v0.23.1), \textsf{MulticoreTSNE} (v0.1), and
	\textsf{FIt-SNE} (v1.1.0).  The first section of the comparison
	addresses packaging and distribution. A properly packaged library is
	easily accessible to users, and developers should easily include it in
	dependency lists of other software packages. The second section of the
	comparison lists the two existing t-SNE approximation schemes. The
	FIt-SNE approximation scheme is required for t-SNE to scale up to
	millions of data points. The final section provides a list of extensions
	and improvements to the standard t-SNE algorithm, many of which can
	produce markedly better visualizations.
}
\begin{center}\small
\newcommand*\rot{\rotatebox{90}}
\renewcommand{\arraystretch}{1.25}

\begin{tabular}{r c c c c c c}
\toprule
\setlength\tabcolsep{6pt}
& \rot{\textsf{scikit-learn}} & \rot{\textsf{MulticoreTSNE}} & \rot{\textsf{FIt-SNE}} & \rot{\textsf{openTSNE}} \\
\toprule
\textsf{PyPI} package & \checkmark & \checkmark & & \checkmark \\
\textsf{conda} package & \checkmark & & & \checkmark \\
Precompiled binaries & \checkmark & & & \checkmark \\
\hline
Barnes-Hut ($\mathcal{O}(N \log N)$) & \checkmark & \checkmark & & \checkmark \\
FIt-SNE ($\mathcal{O}(N)$) & & & \checkmark & \checkmark \\
\hline
Multiscale Gaussian kernels & & & \checkmark & \checkmark \\
Fully-custom affinity kernels & & & & \checkmark \\
Variable degrees of freedom & & & \checkmark & \checkmark \\
Variable exaggeration & & & \checkmark & \checkmark \\
Better initialization & & & \checkmark & \checkmark \\
Automatic learning rate & & & \checkmark & \checkmark \\
Embedding new samples & & & & \checkmark \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
